Software Architectures for Integrating Workflow Engines in Science Gateways

### 2016-05-06 paper.tex
Reverted title to original one
A few fixes in the abstract and introduction

\title{A comprehensive analysis of existing software architectures\\ to integrate workflow engines in science gateways\closednote{Naj}{'what' software architectures? what is their qualifier, adjective?}{Tristan}{There is no qualifier except 'for integrating workflow engines in science gatewats'. They are just software architectures :) The point of the paper is to describe them properly. It's a general formulation to cover any kind of software architecture that is used to integrated workflow engines in science gateways. Does it make sense?}}

@@

\begin{abstract}
  We investigate six software architectures commonly used to integrate
  workflow engines in science gateways. (1) In \emph{tight
    integration}, the workflow engine shares software components with
  the science gateway; (2) In \emph{service invocation}, the engine is
  isolated and invoked through a specific software interface; (3) In
  \emph{task encapsulation}, the engine is wrapped as a computing task
  executed on the infrastructure; (4) In the \emph{pool model}, the
  engine is bundled in an agent that connects to a central pool to
  fetch and execute workflows; (5) In \emph{nested workflows}, the
  engine is integrated as a child process of another engine; (6) In
  \emph{workflow import}, the engine is integrated through workflow
  language conversion. We describe and evaluate these architectures
  with metrics that measure integration complexity, robustness,
  extensibility, scalability and support for meta-workflows and
  fine-grained debugging\closednote{Naj}{Please be precise, what are
    they?}{Tristan}{Done, thanks}. Results provide insights for
  science gateway architects and workflow engine developers. Tight
  integration and task encapsulation are the easiest to integrate and
  are most robust. Extensibility is equivalent in most
  architectures. The pool model is the most scalable one and
  meta-workflows are only available in nested workflows and workflow
  import. \note{Naj}{I find this abstract to be vague, it can be made
    more concise and poignant}
\end{abstract}

@@

Workflow engines are critical for the efficient and transparent
exploitation of distributed infrastructures in the ecosystem of tools
and services offered by science gateways. Several software
architectures can be adopted to integrate workflow engines in science
gateways, however development effort required and the resulting system
vary across different architectures.

This paper describes, provides examples and compares such
architectures, based on system-independent representations of their
main components and interactions. It is informed by our experince in
the development and sustained operation of the CBRAIN~\cite{SHER-14}
and VIP~\cite{GLAT-13} science gateways for medical image analysis
during the past 7 years; as well as by lessons learned from several
science gateway and workflow projects such as
SHIWA\footnote{\url{http://www.shiwa-workflow.eu}}.  In this section
we provide background information and definitions of different
workflow engines, science gateways and infrastructures. In
Section~\ref{sec:architectures}, we describe six architectures within
a consistent framework that underlines the functional interactions
between their main software components. In
Section~\ref{sec:evaluation}, the different architectures are
evaluated with metrics measuring integration complexity, robustness,
extensibility, scalability and other specific features. We finally
compare the architectures and highlight recommendations for science
gateway architects and workflow engine developers.

@@

At the same time, toolboxes have been emerging in various scientific
domains to facilitate the assembly of software components in
consistent ``pipelines''. In neuroimaging, our primary domain of
interest, tools such as Nipype (Neuroimaging in Python, Pipelines and
Interfaces~\cite{gorgolewski2011nipype}), PSOM (Pipeline System for
Octave and Matlab~\cite{bellec2012pipeline}), RPPL~\cite{1174106} and PMP
(Poor Man's Pipeline~\cite{Ad-DabbaghY2006}), SPM (Statistical Parametric
Mapping~\cite{ashburner2011spm}) and FSL (FMRIB Software
Library~\cite{Jenkinson2012782}) provide abstractions and functions to
handle the data and computing flow between processes implemented in a
variety of programming languages. Such tools were interfaced to
computing infrastructures, in particular
clusters\closednote{Tristan}{Reverted from Marc's 'cluster
  scheduler'}{}{}, to execute tasks at a high throughput. Some of
these tools also support advanced features such as provenance tracking
(Nipype, PSOM), \closednote{Naj}{should we not mention PMP which is
  used for CIVET?}{Tristan}{Done, as discussed.}
redundancy detection across analyses to avoid re-computation (PSOM),
and so on. A wide-array of workflows have been implemented using these
pipelines and are now shared across neuroimaging groups
world-wide. These domain-specific engines represent a tremendous
opportunity for science gateways to leverage existing tools and
applications. They nicely complement e-Science engines that are more
oriented towards the exploitation of distributed computing
infrastructures, in particular grids and clouds.

@@

Various science gateways have been developed, including complete
frameworks such as Apache Airavata~\cite{marru2011apache}, the Catania
Science Gateway Framework~\cite{Ardizzone2012} and
WS-PGRADE/gUSE~\cite{Kacsuk2012}. Numerous science gateways were built
using such frameworks~\cite{kacsuk2014science,ardizzone2012decide} or
as standalone systems~\cite{SHER-14,GLAT-13}. Most of these science
gateway frameworks and instances include one or several workflow
engines.

@@

Integration between workflow engines and science gateways varies
across systems. Some science gateways are tailored to a particular
engine while other ones are more general and host applications
executed by different types of engines. Some gateways may not even use any
workflow engine and focus on the execution of simple tasks.

Extensibility is an important property of the integration, as new
workflows are added frequently, different types and versions of
workflow engines may be integrated over time, and different kinds of
infrastructure can be targeted. Scalability is also a crucial concern
for such multi-user systems that manage substantial amounts of data
and execute computationally-intensive analyses. To this end, science
gateways may balance the load among different instances of the same
engine. They may also use advanced task scheduling policies on the
infrastructure, e.g. to ensure fairness among
users~\cite{FERR-14,CPE:CPE3708} or to improve
fault-tolerance~\cite{wilkins2008teragrid,FERR-13}.

Robustness is a highly desirable characteristic as well. Architectures
showing reduced system complexity will greatly facilitate the
implementation effort towards robust interactions which, in turn, have
a positive impact on characteristics such as gateway predictability,
transparence, general reliability and results traceability and
reproducibility; characteristics that are also key to a good user
experience. \closednote{Marc-Etienne}{I find that something is weak in
  this last paragraph. Robustness will help with transparency, sure,
  but it seems only a small facet.  Also, the text seem to timidly
  favor architectures with less components/interactions without much
  explanation... Complex systems can be robust, but at greater
  costs...  Here is one: "Architectures showing reduced system
  complexity will greatly facilitate the implementation effort towards
  robust interactions which, in turn, have a positive impact on
  characteristics such as gateway predictability, transparence,
  general reliability and results traceability and reproducibility;
  characteristics that are also key to a good user
  experience."}{Tristan}{Agreed, fixed as
  suggested.}\note{Tristan}{Work on the this sentence again.}

>>>>

\title{Software architectures to integrate workflow engines in science gateways\closednote{Naj}{'what' software architectures? what is their qualifier, adjective?}{Tristan}{There is no qualifier except 'for integrating workflow engines in science gatewats'. They are just software architectures :) The point of the paper is to describe them properly. It's a general formulation to cover any kind of software architecture that is used to integrated workflow engines in science gateways. Does it make sense?}}

@@

\begin{abstract}
  We investigate six software architectures commonly used to integrate
  workflow engines in science gateways. In \emph{tight integration},
  the workflow engine shares software components with the science
  gateway. In \emph{service invocation}, the engine is isolated and
  invoked through a specific software interface. In \emph{task
    encapsulation}, the engine is wrapped as a computing task executed
  on the infrastructure. In the \emph{pool model}, the engine is
  bundled in an agent that connects to a central pool to fetch and
  execute workflows. In \emph{nested workflows}, the engine is
  integrated as a child process of another engine. In \emph{workflow
    import}, the engine is integrated through workflow language
  conversion. We describe and evaluate these architectures with
  metrics that measure integration complexity, robustness,
  extensibility, scalability and support for meta-workflows and
  fine-grained debugging\closednote{Naj}{Please be precise, what are
    they?}{Tristan}{Done, thanks}. Results provide insights for
  science gateway architects and workflow engine developers. Tight
  integration and task encapsulation are the easiest to integrate and
  are most robust. Extensibility is equivalent in most
  architectures. The pool model is the most scalable one and
  meta-workflows are only available in nested workflows and workflow
  import. \closednote{Naj}{I find this abstract to be vague, it can be made
    more concise and poignant}{Tristan}{Added a description of the
    architectures, following @pbellec's suggestion.}
\end{abstract}

@@

Workflow engines are critical for the efficient and transparent
exploitation of distributed infrastructures in the ecosystem of tools
and services offered by science gateways. Several software
architectures can be adopted to integrate workflow engines in science
gateways, with important consequences on the development effort
required and resulting system.

This paper describes, provides examples and compares such
architectures, based on system-independent representations of their
main components and interactions. It is informed by our experince in
the development and sustained operation of the CBRAIN~\cite{SHER-14}
and VIP~\cite{GLAT-13} science gateways for medical image analysis
during the past 7 years; as well as by lessons learned from several
science gateway and workflow projects such as
SHIWA\footnote{\url{http://www.shiwa-workflow.eu}}.  In this section
we provide background information and definitions of 
workflow engines, science gateways and infrastructures. In
Section~\ref{sec:architectures}, we describe six architectures within
a consistent framework that underlines the functional interactions
between their main software components. In
Section~\ref{sec:evaluation}, the different architectures are
evaluated with metrics measuring integration complexity, robustness,
extensibility, scalability and other specific features. We finally
compare the architectures and highlight recommendations for science
gateway architects and workflow engine developers.

@@

At the same time, toolboxes have been emerging in various scientific
domains to facilitate the assembly of software components in
consistent ``pipelines''. In neuroimaging, our primary domain of
interest, tools such as Nipype (Neuroimaging in Python, Pipelines and
Interfaces~\cite{gorgolewski2011nipype}), PSOM (Pipeline System for
Octave and Matlab~\cite{bellec2012pipeline}), RPPL~\cite{1174106} and
PMP (Poor Man's Pipeline~\cite{Ad-DabbaghY2006}), SPM (Statistical
Parametric Mapping~\cite{ashburner2011spm}) and FSL (FMRIB Software
Library~\cite{Jenkinson2012782}) provide abstractions and functions to
handle the data and computing flow between processes implemented in a
variety of programming languages. Such tools were interfaced to
computing infrastructures, in particular
clusters\closednote{Tristan}{Reverted from Marc's 'cluster
  scheduler'}{}{}, to execute tasks at a high throughput. Some of
these tools also support advanced features such as provenance
tracking, \closednote{Naj}{should we not mention PMP which is used for
  CIVET?}{Tristan}{Done, as discussed.}  redundancy detection across
analyses to avoid re-computation, and so on. A wide-array of
workflows have been implemented using these pipelines and are now
shared across neuroimaging groups world-wide. These domain-specific
engines represent a tremendous opportunity for science gateways to
leverage existing tools and applications. They nicely complement
e-Science engines that are more oriented towards the exploitation of
distributed computing infrastructures, in particular grids and clouds.

@@

Various science gateways have been developed, including complete
frameworks such as Apache Airavata~\cite{marru2011apache}, the Catania
Science Gateway Framework~\cite{Ardizzone2012} and
WS-PGRADE/gUSE~\cite{Kacsuk2012}. Numerous science gateways were built
using such frameworks~\cite{kacsuk2014science,ardizzone2012decide} or
as standalone systems~\cite{SHER-14,GLAT-13}. Most of these frameworks
and instances include one or several workflow engines.

@@

Integration between workflow engines and science gateways varies
across systems. Some science gateways are tailored to a particular
engine while other ones are more general and host applications
executed by different types of engines. Some gateways may not even use any
workflow engine and instead focus on the execution of simple tasks.

Extensibility is an important property of the integration. New
workflows are added frequently, different types and versions of
workflow engines may be integrated over time, and different kinds of
infrastructure can be targeted.

Scalability is also a crucial concern for such multi-user,
high-throughput systems. To this end, science gateways may balance the
load among different instances of the same engine, and use advanced
task scheduling policies on the infrastructure to improve performance,
fault-tolerance and fairness among users.

Robustness is highly desirable as well since it is key to a
good user experience. Simple architectures facilitate the
implementation effort towards robust interactions which, in turn, have
a positive impact on characteristics such as gateway predictability,
transparency, reliability, traceability and
reproducibility. \closednote{Marc-Etienne}{I find that something is
  weak in this last paragraph. Robustness will help with transparency,
  sure, but it seems only a small facet.  Also, the text seem to
  timidly favor architectures with less components/interactions
  without much explanation... Complex systems can be robust, but at
  greater costs...  Here is one: "Architectures showing reduced system
  complexity will greatly facilitate the implementation effort towards
  robust interactions which, in turn, have a positive impact on
  characteristics such as gateway predictability, transparence,
  general reliability and results traceability and reproducibility;
  characteristics that are also key to a good user
  experience."}{Tristan}{Agreed, fixed as
  suggested.}\closednote{Tristan}{Work on the this sentence again.}{Tristan}{done.}
